Categorical Foundations of E-Graphs With Binding

### 2025-07-11 POPL2026/paper/CategoricalFoundationsOfEGraphsWithBinding.tex
some polish

\section{Conclusion}
In this work we introduced e-graphs with bindings formalised as morphisms of a closed symmetric monoidal $\catname{SLat}$-category.
In doing so, we are able to leverage its internal language to provide variable binding in $\lambda$-calculus-like languages by metatheoretic $\lambda$-abstraction.
We then provided a combinatorial representation for such morphisms in terms of closed e-hypergraphs with a suitable notion of sound and complete DPO rewriting, making it possible for a closed e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
We argued for the utility of our techniques in the representation of programming languages (which inherently require variable binding) as e-graphs, contrasting with the more general-purpose slotted e-graphs, and advocating for the benefits of string-diagrammatic syntax.
In future, we hope to implement an efficient version of e-graphs for hypergraphs as a data structure using DPO rewriting as its foundation to implement e-graph operations, with the theory we have developed here acting as a roadmap.

>>>>

\section{Conclusion}
In this work we introduced e-graphs with bindings derived as morphisms of a closed symmetric monoidal $\catname{SLat}$-category.
In doing so, we are able to leverage its internal language to provide variable binding in $\lambda$-calculus-like languages by metatheoretic $\lambda$-abstraction.
We then provided a combinatorial representation for such morphisms in terms of closed e-hypergraphs with a suitable notion of sound and complete DPO rewriting, making it possible for a closed e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
We argued for the utility of our techniques in the representation of programming languages (which inherently require variable binding) as e-graphs, contrasting with the more general-purpose slotted e-graphs, and advocating for the benefits of string-diagrammatic syntax, in particular in the context of programming languages.
In the future, we hope to implement an efficient version of e-graphs for hypergraphs as a data structure using DPO rewriting as its foundation to implement e-graph operations, with the theory we have developed here acting as a roadmap.

### 2025-07-11 POPL2026/paper/sections/categorical_semantics.tex
some polish

To aid reasoning, we introduce a new language of string diagrams for closed symmetric monoidal $\catname{SLat}$-categories, using a hierarchical \enquote{dashed box} structure to capture the join operation on morphisms (as was done by~\citet{tiurin2025equivalencehypergraphsdporewriting}) as well as another hierarchical \enquote{rounded box} for $\lambda$-abstraction.
\autoref{fig:egraph-strings} displays the generators of this language, which is the usual string diagrammatic syntax~\cite{Selinger_2010} plus the syntax for $\lambda$-abstraction as established by~\citet{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice enrichment as introduced by~\citet{tiurin2025equivalencehypergraphsdporewriting}; the first component denotes an empty diagram.

>>>>

To aid reasoning, we introduce a new language of string diagrams for closed symmetric monoidal $\catname{SLat}$-categories, using a hierarchical \enquote{dashed box} structure to capture the join operation on morphisms (as was done by~\citet{tiurin2025equivalencehypergraphsdporewriting}) as well as another hierarchical \enquote{rounded box} for $\lambda$-abstraction.
\autoref{fig:egraph-strings} displays the generators of this language, which is the usual string-diagrammatic syntax~\cite{Selinger_2010} plus the syntax for $\lambda$-abstraction as established by~\citet{ghica2024stringdiagramslambdacalculifunctional} and syntax for semilattice enrichment as introduced by~\citet{tiurin2025equivalencehypergraphsdporewriting}; the first component denotes an empty diagram.

### 2025-07-11 POPL2026/paper/sections/introduction.tex
some polish

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach hides the bureacracy in complex operations and reveals insights sometimes obscured by traditional one-dimensional symbolic notation.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition.
The explicit occurrences of variable names in the conventional approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices instead when performing $\beta$-reduction would only push the problem elsewhere, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.
Recent work on \emph{slotted} e-graphs~\cite{slotted-egraphs} uses nominal techniques to address the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by incorporating variables as first-class data of an e-graph and making binding a built-in operation.
They also solved the problem of explicit variable names, guaranteeing freshness inside binders, explicitly differentiating between bound and free variables.
An example of a slotted e-graph is given in~\autoref{fig:slotted}.

@@

\begin{figure*}
	\begin{subfigure}{\linewidth}
		\[
			% \hspace{1.3cm}
			\adjustbox{width=\textwidth}{
				\tikzfig{figures/e-graph-substitution}
			}
		\]
		% \captionsetup{skip=0pt}
		\subcaption{E-graph explicit substitution example.}
		\label{fig:e-graph-substitution}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\[
			% \hspace{1.3cm}
			\adjustbox{width=\textwidth}{
				\tikzfig{figures/e-string-substitution-2}
			}
		\]
		\subcaption{String diagrammatic substitution example.}\label{fig:e-string-substitution}
	\end{subfigure}
\end{figure*}
%   \begin{figure*}
%     \[
%         % \hspace{1.3cm}
%         \adjustbox{width=\textwidth}{
%         \tikzfig{figures/e-string-substitution-2}
%         }
%     \]
% 	\captionsetup{skip=0pt}
%     \caption{String diagrammatic substitution example.}
%     \label{fig:e-string-substitution}
%   \end{figure*}

@@

\citet{tiurin2025equivalencehypergraphsdporewriting} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free Cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

>>>>

String diagrams provide a topological framework for understanding monoidal categories, representing objects as \emph{strings} and morphisms as \emph{nodes}, where input strings flow in and output strings flow out.
In this article, our string diagrams are oriented left-to-right, bottom-to-top to make the pictures look more similar to those commonly drawn for e-graphs.
The structure of symmetric monoidal categories is absorbed into the graphical syntax, with the composition and tensoring represented by juxtaposition, and symmetries as weaving of strings.
This visual approach hides the bureacracy in complex operations and reveals insights sometimes obscured by traditional one-dimensional symbolic notation.
Originating from notations by~\citet{hotzsd} (for automata) and~\citet{penrose1984spinors} (for tensor networks), extensions of string-diagrammatic calculi exist for various forms of monoidal categories and even bicategories~\cite{Selinger_2010}, and provide a mathematically rigorous reasoning toolkit~\cite{joyal_geometry_1991}.

@@

Using string diagrams, $\beta$-reduction is simply the re-wiring of the diagram where an application node meets an abstraction, and explicit substitution simplifies into mere composition.
The explicit occurrences of variable names in the conventional approach raise obvious questions of scope, binding and explicit substitution, whereas the use of terms inside the explicit substitution nodes also raises many technical difficulties which makes proving various mathematical properties essentially intractable.
Additionally, explicit substitutions need to be propagated recursively throughout the graph, which adds another layer of technical and conceptual overhead.
Using de Bruijn indices instead when performing $\beta$-reduction would only push the problem elsewhere, as it greatly increases the complexity of representing the reduction as a rewrite rule, and special shifting operators need to be introduced on the right-hand side of a rewrite rule, which is a meta-syntactic operation~\cite{koehler2022sketchguided}.

Recent work on \emph{slotted} e-graphs~\cite{slotted-egraphs} uses nominal techniques to address the sharing issues when representing terms that are equivalent up to renaming of variables (even free ones) by incorporating variables as first-class data of an e-graph and making binding a built-in operation.
They also solved the problem of explicit variable names, guaranteeing freshness inside binders, explicitly differentiating between bound and free variables.
An example of a slotted e-graph is given in~\autoref{fig:slotted}.

@@

\begin{figure*}
	\begin{subfigure}{\linewidth}
		\[
			% \hspace{1.3cm}
			\adjustbox{width=\textwidth}{
				\tikzfig{figures/e-graph-substitution}
			}
		\]
		% \captionsetup{skip=0pt}
		\subcaption{E-graph explicit substitution example.}
		\label{fig:e-graph-substitution}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\[
			% \hspace{1.3cm}
			\adjustbox{width=\textwidth}{
				\tikzfig{figures/e-string-substitution-2}
			}
		\]
		\subcaption{String-diagrammatic substitution example.}\label{fig:e-string-substitution}
	\end{subfigure}
	\caption{Conventional e-graphs vs.\ categorical e-graphs with binding.}
\end{figure*}
%   \begin{figure*}
%     \[
%         % \hspace{1.3cm}
%         \adjustbox{width=\textwidth}{
%         \tikzfig{figures/e-string-substitution-2}
%         }
%     \]
% 	\captionsetup{skip=0pt}
%     \caption{String-diagrammatic substitution example.}
%     \label{fig:e-string-substitution}
%   \end{figure*}

@@

Our contribution is the construction of a form of e-graphs that support binding, developed through a principled categorical approach.
Specifically, we extend a prior categorical semantics for e-graphs without binding by incorporating a monoidal closed structure.
\citet{tiurin2025equivalencehypergraphsdporewriting} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free Cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.
